HO unification from object language to meta language

### 2024-04-01 paper/main.tex
adapt tex to print beta links
Update maybebeta compilation section

\subsection{Compilation}
\marginpar{manca beta normal in entrata}

@@

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
we modify the term compiler by putting fresh unification variables
in their place: the problematic term is moved from 
from  \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
of the problem \foUnifPb above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

@@

A very basic example of \linketa deduplication, is given below:
% make test ONLY=7002 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),\lambda x.(Y\appsep x)}}}
  {{{A,C}}}
  {{{X,B,1},
    {Y,D,1}}}
  {{{,A,\lambda x.B_{x}},
    {,C,\lambda x.D_{x}}}}

@@

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective} (merging the two mappings
for $s$ would break \cref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{x,E_{x},\lambda y.C_{x y}},
    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}
\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

@@

Detection of \maybebeta is quite simple to implement in the compiler, it is 
sufficient to

Dire che maybebeta is simple to detect e che maybe eta funziona bene and su 
termini che non sono il llambda

\begin{corollary}
  Right hand side is $X_{scope} t_1 \dots t_n$ and $t_1$ is in $scope$ or
  $t_1$ is not a name
\end{corollary}

@@

A term is \maybebeta if it has the shape \elpiIn{fapp[fuva A|Ag]} and
\elpiIn{distinct Ag} does not hold. In that case, \elpiIn{Ag} is split in two
sublist \elpiIn{Pf} and \elpiIn{Extra} such that former is the longest prefix of
\elpiIn{Ag} such that \elpiIn{distinct Pf} holds. \elpiIn{Extra} is the list
such that \elpiIn{append Pf Extra Ag}. Next important step is to compile
recursively the terms of these lists and allocate a memory adress \elpiIn{B}
from the substitution in order to map the \Fo variable \elpiIn{fuva A} to
the \Ho variable \elpiIn{uva B}. The \linkbeta to return in the end is given
by the term \elpiIn{Beta = app[uva B Scope1 | Extra1]} constituting the \rhs,
and a fresh variable \elpiIn{C} having in scope all the free variables occurring
in \elpiIn{Beta} (this is \lhs). We point out that the \rhs is intentionally
built as an \elpiIn{uva} where \elpiIn{Extra1} are not in scope, since by
invariant, we want all the variables appearing in \Ho to be in \llambda.

>>>>

\subsection{Compilation}
\label{sec:compilation}
\marginpar{manca beta normal in entrata}

@@

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
we modify the term compiler by putting fresh unification variables
in their place: the problematic term is moved from 
from  \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
of the problem \foUnifPb above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{\eta,,A,\lambda x.B_x}}}

@@

A very basic example of \linketa deduplication, is given below:
% make test ONLY=7002 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),\lambda x.(Y\appsep x)}}}
  {{{A,C}}}
  {{{X,B,1},
    {Y,D,1}}}
  {{{\eta,,A,\lambda x.B_{x}},
    {\eta,,C,\lambda x.D_{x}}}}

@@

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective} (merging the two mappings
for $s$ would break \cref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

@@

Detection of \maybebeta is quite simple to implement in the compiler, since it
is sufficient to capture applications with flexible head and argument that
are not in \llambda.

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the compiled
list of terms obtained from \elpiIn{Pf} and \elpiIn{Extra1} as arguments. The
variable \elpiIn{B} returned has compiled terms is a fresh variable having in
scope all the free variables used in \rhs.

\begin{invariant}
  The \rhs of a \linkbeta has the shape \elpiIn{app[uva N S|L]}.
\end{invariant}

\begin{corollary}
  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then \elpiIn{L} is not
  empty.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, if \elpiIn{L} is a empty list, then the original \Fo{}
  term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a list of 
  disting names. This case is however captured by rule~\ref{rule:complam} (from
  \cref{sec:compilation}) and no \linkbeta is produced: a contradiction.
\end{proof}

\begin{corollary}
  Let \elpiIn{app[uva N S|L]} be the \rhs of a \linkbeta, then the 
  first argument $t$ in \elpiIn{L} either appears in \elpiIn{S} or it is 
  not a name.
\end{corollary}

\begin{proof}[Proof sketch]
  By construction, the lists \elpiIn{S} and \elpiIn{L} are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  \elpiIn{S} is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, either the first element of
  \elpiIn{L} is a name appearing in \elpiIn{S} or it a term with a constructor
  of \elpiIn{tm} as functor.  
\end{proof}

TODO: Dire che maybe eta fa il detect anche su 
termini che non sono il llambda, oppure dirlo in section of maybeeta + dare un
esempio?
